### 2015-01-05 document/article.tex
add document

\documentclass[12pt]{article}
\usepackage{fullpage,graphicx,psfrag,amsmath,amsfonts,verbatim}
\usepackage{listings, color}
\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}
\usepackage[scaled]{beramono}
\usepackage{courier, natbib}
\usepackage[usenames,dvipsnames]{xcolor}
\usepackage{hyperref, adjustbox}
\usepackage[small,bf]{caption}
\usepackage{float}
\usepackage{booktabs}

\newcommand{\subf}[2]{%
  {\small\begin{tabular}[t]{@{}c@{}}
  #1\\#2
  \end{tabular}}%
}

\lstset{
  language=Python,
  backgroundcolor=\color{white},
  breaklines=true,
  captionpos=b,
  commentstyle=\color{mygreen},
  showstringspaces=false,
  formfeed=newpage,
  keepspaces=true,
  stringstyle=\color{mymauve},
  keywordstyle=\color{blue},
  numbers=left,
  tabsize=4,
  numberstyle=\footnotesize,
  basicstyle=\footnotesize,
  morekeywords={models, lambda, forms, True, False, self}
}

\newcommand{\code}[2] {
  \hrulefill
  \subsection*{#1}
  \lstinputlisting{#2}
  \vspace{2em}
}

\newcommand{\tab}[1]{\hspace{.2\textwidth}\rlap{#1}}

\input defs.tex

\title{\texttt{CompleteThat}: A python package for  low-rank matrix completion}
\author{Joshua Edgerton \and Esteban Fajardo}
\date{January 5, 2014}

\begin{document}

\maketitle

\begin{abstract}
We have developed a Python package (\texttt{CompleteThat}) to perform low rank matrix completion. Given a low rank matrix with partial entries we implemented different methods to solve the matrix completion problem:  First, for ``small'' problems we implemented two memory-based algorithms  following the work by Tanner and Wei~\cite{Tanner:2014} to find, given a number $r$, the matrix with rank $r$ that best fits the data using the Frobenius norm. Second, when the problem does not fit into memory, we implemented a memory-fitting algorithm (stochastic gradient descent) following the approach in Zhang~\cite{zhang:2004} and Bottou~\cite{bottou:2012}. The package is available at \url{https://pypi.python.org/pypi/completethat/0.1dev}. 
\end{abstract}

\newpage
\tableofcontents
\newpage

\section{Introduction}
Matrix factorization has recently seen a large growth in popularity within the mathematics, statistics, and computer science communities as industry continues to apply machine learning techniques on a wide array of problems and scenarios. For our convex optimization project, we decided to take some of the more interesting and applicable topics and algorithms of our class and develop a python package to implement them. We briefly discuss the theory behind matrix factorizing, the models and approaches we choose to use and the algorithms for the optimization. Later, we walk through two case studies illustrating the usefulness and applicability in the context of image process and a recommendation system for Yahoo music and movie data.

\section{Matrix Completion Problem}

\begin{quote}
Consider the problem
    \begin{equation}
    \begin{array}{ll}
    \mbox{minimize}   & (1/2)\|Ax-b\|_2^2 + \lambda \|x\|_1 \\
    \mbox{subject to} & 0 \preceq x \preceq \ones \\
    & \|x\|_2 \leq 1,
    \end{array}
    \label{eq:original_problem}
    \end{equation}
where $x \in \reals^n$ is the optimization variable, and $A \in \reals^{m
\times n}$, $b \in \reals^m$, and $\lambda > 0$ are problem data.
\end{quote}

\section{Implementation}
\texttt{CompleteThat} is a python package that solves the low rank matrix completion problem. Given a low rank matrix with partial entries the package solves an optimization problem to estimate the missing entries. We allow the user to choose between several optimization algorithms including two in memory based algorithms, alternating steepest descent and scaled alternating steepest descent, and one out of memory fitting procedure using stochastic gradient descent. We use extensively the numerical libraries \texttt{spicy} and \texttt{numpy}, and the current implementation includes two classes, \texttt{MatrixCompletion} for in memory based algorithms and \texttt{MatrixCompletionBD} for the memory fitting procedure. The package is available worldwide and can be
downloaded from \url{https://pypi.python.org/pypi/completethat/0.1dev}.

\subsection*{CVX}
Problem (\ref{eq:original_problem}) can be easily solved directly using CVX, a package for specifying and solving convex programs~\cite{cvx}, ~\cite{gb08}, with the following code:

\begin{verbatim}
index = find(~isnan(M));
cvx_begin
    variable X(size(M));
    minimize norm_nuc(X)
    % s.t.
    X(index) == M(index);
cvx_end
\end{verbatim}

Where M is a MATLAB matrix with nan on the missing entries and \texttt{X(index) == M(index)} corresponds to the $P_{\Omega}(X) = P_{\Omega}(M)$ restriction. However, the computation is very slow and for any matrix greater than 100x100 the computational time is too high. Since this is clearly unsatisfactory for practical purposes, specific algorithms were developed and used on the package.

\subsection*{Algorithms}
\subsubsection*{Low rank matrix completion by alternating steepest descent methods}
\subsubsection*{Low rank matrix completion by stochastic gradient descent}

\subsection*{Code Example}
For matrices that fit into memory use the \texttt{MatrixCompletion} module. Otherwise, use \texttt{MatrixCompletionBD} module.

\subsubsection*{MatrixCompletion}
Given a \texttt{numpy} matrix $M$ with \texttt{numpy.nan} on the missing entries, the matrix completion problem can be solved (using ASD) as:
\begin{verbatim}
>>> from completethat import MatrixCompletion
>>> problem = MatrixCompletion(M)
>>> problem.complete_it("ASD")
>>> X = problem.get_matrix() #Desired matrix
>>> out_info = problem.get_out() #Extra information
\end{verbatim}

\subsubsection*{MatrixCompletionBD}
Given a csv file with the input data, the matrix completion problem can be solved as:

\begin{verbatim}
>>> from completethat import MatrixCompletionBD
>>> problem = MatrixCompletionBD('input_data.txt')
>>> problem.train_sgd(dimension=6,init_step_size=.01,min_step=.000001, 
       reltol=.001,rand_init_scale=10, maxiter=1000,
       batch_size_sgd=50000,shuffle=True)
>>> problem.validate_sgd('test_data.txt')
>>> problem.save_model()
\end{verbatim}

\section{Case Studies}
\subsection*{Yahoo movies and music reviews database}
The Yahoo music and movie data sets are publicly available datasets published by Yahoo for use to the public for recreational and research purposes. The music training data set is a small 4 mb pipe-delimited file consisting of roughly 210,000 user-movie ratings. The test set has roughly 10,000 records. There are two ratings schemas available, one on a 5-point scale and the other on a 13-point scale and we arbitrarily chose to use the 5-point scale. The Yahoo music data is a larger 2.2 gb file of 115 million user-song ratings. The ratings are on a 100 point scale.

We look to evaluate the performance of our algorithms on these two datasets, comparing the various algorithms implemented versus a naive benchmark estimate where we used the mean of the training set as our estimator for all records in the test set. Then we compared how well the algorithms performed versus the benchmark.

The benchmark for the yahoo music data is 4.088. This is the mean value of the ratings in the training set so we will compute the MSE on the test set using the benchmark which yields an RMSE of 1.10. A quick run of Mahout yields 1.17 and CompleteThat's sgd model yielded a 1.19. So our model yielded a pretty similar RMSE as the MAHOUT package. Note that for both of these instances we used a factorization of rank equal to seven and the Complete that sgd algorithm ran slightly faster than Mahout's (.537 minutes vs 1.4 minutes).

\begin{table}[h]
\centering
\resizebox{0.75\textwidth}{!}{%
\begin{tabular}{@{}ccccc@{}}
\toprule
\multicolumn{1}{l}{\textbf{Step size}} & \multicolumn{1}{l}{\textbf{Iterations}} & \multicolumn{1}{l}{\textbf{Minutes}} & \multicolumn{1}{l}{\textbf{Train MSE}} & \multicolumn{1}{l}{\textbf{Test RMSE}} \\ \midrule
1 & 8 & 0.585 & 1.411 & 1.757 \\
2 & 8 & 0.598 & 1.304 & 1.550 \\
3 & 8 & 0.628 & 1.235 & 1.483 \\
4 & 8 & 0.617 & 1.180 & 1.397 \\
5 & 7 & 0.538 & 1.136 & 1.351 \\
6 & 7 & 0.534 & 1.100 & 1.244 \\
7 & 7 & 0.534 & 1.071 & 1.194 \\
8 & 7 & 0.541 & 1.046 & 1.127 \\
9 & 7 & 0.537 & 1.027 & 1.128 \\
10 & 7 & 0.529 & 1.010 & 1.102 \\
15 & 7 & 0.551 & 0.971 & 1.063 \\
20 & 7 & 0.540 & 0.976 & 1.075 \\
30 & 8 & 0.614 & 1.045 & 1.359 \\ \bottomrule
\end{tabular}
}
\caption{RMSE vs Step Size vs Rank}
\end{table}

\subsection*{Columbia University images}

\newpage
\begin{figure}[H]
\makebox[\textwidth][c]
{
\begin{tabular}{ccc}
\includegraphics[width=65mm]{figures/columbia1.png}
&
\includegraphics[width=65mm]{figures/columbia2.png}

&
\includegraphics[width=65mm]{figures/columbia3.png}

\\

\includegraphics[width=65mm]{figures/columbia4.png}
&
\includegraphics[width=65mm]{figures/columbia5.png}
&
\includegraphics[width=65mm]{figures/columbia6.png}
\\

\includegraphics[width=65mm]{figures/columbia7.png}
&
\includegraphics[width=65mm]{figures/columbia8.png}
&
\includegraphics[width=65mm]{figures/columbia9.png}
\\
\end{tabular}

}

\caption{ASD method applied three different Columbia University images}
\end{figure}

\newpage

\section{Future Work}

We have plans for various improvements as we continue working with the \texttt{CompleteThat} python package. Overall, we would like to test the software and integrate automated testing cases into the software developing cycle. Also, documentation for the package and its functions, including examples, is high in our priority list. In addition, we hope to incorporate more rigorous error checking and make   available more customization to the user.

For the memory-based algorithms it is well know that noise in the data introduces overfitting on the estimated matrix. Following the approach taken by Mazumder et al.~\cite{mazumder:2010}  where they solve the same objetive function as the problems above (using the Frobenious norm) but introducing the nuclear norm as regularizer to account for overfitting, we would introduce a regularization option into the matrix completion procedure.

For the stochastic gradient descent method, we plan to add a lambda penalty feature for combatting overfitting as well as considering options for more advanced versions of our basic latent factor model as demonstrated by Koren~\cite{koren:2008} in his paper on the prize-winning Netflix models. 
For all algorithms, we would like to explore and research various hardware and software optimization techniques for faster code.

Finally, given the similarities between matrix completion and robust principal component analysis one further extension to the functionality of the package could be implementing the robust-PCA procedure outlined by Cand\'es~\cite{candes:2011}, et al.

\newpage
\bibliographystyle{alpha}
\bibliography{references}

\newpage
\section{Source code}
On the whole, the package directory structure looks like the Figure~\ref{fig:code_scheme}. In the following we present the source code for the package.
\begin{figure}[h!]
  \centering
    \includegraphics[width=0.8\textwidth]{./figures/code_scheme.pdf}
    \caption{CompleteThat structure}
    \label{fig:code_scheme}
\end{figure}

\code{completethat/matrix\_completion.py}{../code/completethat/matrix_completion.py}
\code{completethat/matrix\_\_init\_\_.py}{../code/completethat/__init__.py}
\code{completethat/matrix\_\_init\_\_.py}{../code/setup.py}

\newpage

\paragraph{Punctuation in equations.}
An equation is part of a sentence, and you may need to include a comma or a period
at the end of an equation as a result, whether or not you are using inline or
display math style. For example:
\begin{quote}
    We next discuss how to solve the problem
    \[
        \begin{array}{ll}
            \mbox{minimize} & (1/2)\|Ax - b\|_2^2,
        \end{array}
    \]
    where $x \in \reals^n$ is the optimization variable.
\end{quote}

\paragraph{Don't start a sentence with a symbol.}
This hurts readability:
\begin{quote}
Bad: $f$ is smooth.\\
Good: The function $f$ is smooth.

Bad: $x^n - a$ has $n$ distinct zeros. \\
Good: The polynomial $x^n - a$ has $n$ distinct zeros.
\end{quote}
Similarly, don't start a sentence with a reference.

\paragraph{Use words to separate symbols in different formulas.}
If it might confuse the reader visually or in the actual
meaning of the sentence, use words to break apart formulas:
\begin{quote}
Bad: The sequences $x_1, x_2, \dots$, $y_1, y_2, \dots$ are Cauchy. \\
OK: The sequences $x_1, x_2, \dots,$ and $y_1, y_2, \dots,$ are Cauchy. \\
Good: The sequences $(x_i)$ and $(y_i)$ are Cauchy.

OK: The image of $S$ under $f$, $f(S) = \{ x \mid x \in S \}$, is convex. \\
Good: The image of $S$ under $f$, given by $f(S) = \{ x \mid x \in S \}$, is convex.
\end{quote}
Do not insert superfluous words if the meaning is clear.
\begin{quote}
Good: Consider the function $f + g + h$, where $f : \reals^n \to
\reals$, $g : \reals^m \to \reals$, and $h : \reals^p \to \symm^n$ are closed
proper convex.
\end{quote}

\paragraph{References.}
For internal references, use the \texttt{label} and \texttt{ref} commands.
\emph{Never} number or refer to an entity using a specific number,
as in ``Table~3.''
Refer to ``Table~\ref{t-ex1}'', ``\eqref{e-opt-prob}'' (an equation), and so on.
Only number equations that are important and should be emphasized or that you
specifically refer to elsewhere. Do not simply number all equations, and do not
number them randomly. A numbered equation draws the reader's attention, so it
should be used relatively sparingly. Use \verb+\S+ for section references, as in
\S\ref{s-content}.

For external references, you must use Bib\TeX. Use references like nouns or 
footnotes:
\begin{quote}
Good: One useful reference is \cite{BoV:04}. \\
Good: Interested readers may refer to Boyd and Vandenberghe \cite{BoV:04}.
\end{quote}

Your Bib\TeX\ source must be correct.  Unfortunately, many systems that export
Bib\TeX\ entries export very poor ones.  Bib\TeX\ ignores the \texttt{*.bib}
file's capitalization for articles (but not books). To force capitalization,
you must wrap the letter with curly braces.  The hyphen (-), en dash (--), and
em dash (---) are distinct punctuation marks that serve different purposes.
(And the minus sign is different from all three.) When specifying a page range
in references, use the en dash. For an example of escaping and using the en
dash, look at our Bib\TeX\ entry for \cite{Tref:2008}.

\paragraph{Do not italicize English in math mode.}
Mathematical symbols should be typeset in math mode:
write $Ax=b$, not Ax=b. This said, subscripts or
superscripts that derive from English (or any human language) should not be
italicized. For example, write $f_\mathrm{best}$, not $f_{best}$. The exception
is subscripts based on a single letter: refer
to a point that is the center of some set as $x_c$, not $x_{\mathrm{c}}$.
Similarly, use commands for special functions: use $\sin(x)$,
$\log(x)$, and $\exp(x)$, not $sin(x)$, $log(x)$, or $exp(x)$.

A really heinous example would be the following:
\begin{quote}
Consider the problem
\[
\begin{array}{ll}
minimize & f(Ax - b) \\
\end{array}
\]
where x is the optimization variable and A and b are problem data.
\end{quote}

\paragraph{Spacing.}
A blank line ends a paragraph. You shouldn't leave a blank line between an
equation and the following text unless you intend the equation to end the
paragraph. Write:
\begin{quote}
\begin{verbatim}
The image of $S$ under $f$,
\[
    f(S) = \{ f(x) \mid x \in S \},
\]
is convex.
\end{verbatim}
\end{quote}
Inserting extra blank lines before \verb+\[+ or after \verb+\]+ will result in
bad typesetting. However, the following is fine, since a new paragraph is called for:
\begin{quote}
\begin{verbatim}
The image of $S$ under $f$ is defined as
\[
    f(S) = \{ f(x) \mid x \in S \}.
\]

We now turn to a different topic.
\end{verbatim}
\end{quote}
Using a tab before \verb+f(S)+ inside the equation environment is
optional, but helps readability in the source.  Similar rules should be
followed for other environments like \texttt{quote}; see the source of this
document.

\paragraph{Use of notation and jargon.}
The correct and appropriate use of notation and jargon takes time to master.
The following should give you a sense of what to think about:

\begin{itemize}
\item Don't use the same notation for two different things. Conversely, be consistent
about notation for the same thing mentioned in two places: don't
say ``$A_j$ for $1 \leq j \leq n$'' in one place and ``$A_i$ for $i=1,\ldots,n$'' 
in another, or use two different indices in summations over the same ranges. 
Formally, the $i$ and $j$ are dummy variables with only that phrase as scope,
so these are technically correct, but it will confuse the reader.

\item It can be useful to choose names for indices so, for example, $i$
always varies from 1 to $m$ and $j$ always varies from 1 to $n$ (when
referencing something like the rows and columns of an $m \times n$ matrix).

\item Define all symbols before or near to where you use them.  A good rule is that
when you first use something, say, $X$, you should either define it immediately, or 
within the paragraph.   There are a few cases where it is acceptable to define it 
later, but you must say this explicitly, as in ``$X$ is the matrix of activation
levels, which we define below.''

\item A symbol like $f$ refers to a function, while $f(x)$ refers to a
function evaluated at a given point. Avoid sloppy writing like ``The function
$f(x)$ is convex.''
So-called `anonymous' functions defined inline are an exception to this rule,
as in ``the function $x^2 \cos x$ is a counterexample,'' though this should
be used only for simple functions.

\item Use typographic naming conventions like capital letters for sets, Greek
letters for dual variables, \emph{etc}.

\item Try to use mnemonic notation, so $x_c$ for a center point,
$c$ for a cost vector, $S$ for a generic set, $C$ for a convex set, $B$
for a ball, and so on. Note that there are conventions about
the use of different letters in mathematics that you should not ignore:
$i$, $j$, and $k$ are usually used for indices, $x$ through $z$ for
variables, $a$ through $d$ for constants, \emph{etc}.

\item Don't use symbols like $\forall$, $\exists$, and $\Rightarrow$; use the
corresponding words. These symbols are usually appropriate only in
formal logic.

\item Don't overuse subscripts. If you do not need to assign a subscript
to something, don't. For example, if you begin by defining a set as
$X = \{x_1, \dots, x_n\}$, then it is going to get annoying to refer to subsets
of $X$, since you will need double subscripts. If possible, it would be better
to not name the elements of $X$ and only refer to them when necessary, just
as, say, $x$ and $y$.

Beware especially of any topic in graph theory when it comes to this rule;
there are often \emph{many} ways to refer to relevant objects, and finding
reasonable notation can often get one half the way to getting a handle on the
problem itself.

\item Don't assign symbols to concepts that you never refer to, or can easily
refer to without:
\begin{quote}
Bad: Let $X$ be a compact subset of a space $Y$. If $f$ is a continuous
real-valued function over $X$, it has a minimum over $X$. \\
Good: A continuous real-valued function has a minimum over a compact set.
\end{quote}
Similarly, do not say ``The solution $x^\star$ is unique'' if you never need
to refer to $x^\star$ again; simply say that the solution is unique.
When you say ``The solution $x^\star$ is unique'' you are both stating a fact
\emph{and} entering the symbol $x^\star$ into the paper's symbol table and
the reader's working memory.
\end{itemize}

\paragraph*{Sections.}
Use structures like \texttt{section}, \texttt{subsection},
\texttt{subsubsection}, and \texttt{paragraph} to organize the exposition;
never insert manual line breaks or page breaks for this purpose. In particular,
do not use line breaks to separate different topics. Do not use symbols in
article titles and ideally not even in section headers.

\paragraph{Use capitalization consistently.}
For section headers, capitalize the first letter, proper nouns, and lowercase
everything else, as in this document.  For references, write Figure 1, Algorithm 3,
Theorem 4, and so on.

\paragraph{Use the right commands.}
There are certain special commands in \LaTeX\ for notation that you otherwise
might attempt to write in an ad-hoc manner. If you do the latter, the typesetting
will be inferior. A few common examples:
\begin{itemize}
\item Norms are written $\|x\|$, not $||x||$.

\item For set-builder notation, use
$\{ x \in \reals \mid x \geq 0 \}$, not $\{x \in \reals | x \geq 0 \}$.

\item Make sure to use \verb+\left+ and \verb+\right+ when wrapping
taller expressions with parentheses, curly braces, brackets, and so on.
\begin{quote}
Bad: Let
\[
    x = \argmin_u (f(u) + \frac{1}{2}\|u - z\|_2^2).
\]
Good: Let
\[
    x = \argmin_u \left(f(u) + \frac{1}{2}\|u - z\|_2^2\right).
\]
\end{quote}

\item Use \verb+\ldots+ (lower dots) when the dots are surrounded by commas and
\verb+\cdots+ (center dots) when surrounded by other objects that have full
height, as in $x_1, x_2, \ldots, x_n$ and $x_1 + x_2 + \cdots + x_n$.
\end{itemize}
Similarly, make sure to use our conventions for standard mathematical objects
for this class; for example, use $\reals$, not $\mathbb{R}$, to denote the set
of real numbers. Use the \verb+\argmin+ and \verb+\argmax+ commands (as shown
above); do not write `arg min', since `argmin' is a single mathematical
operator.  These and other definitions are provided in \verb+defs.tex+.

\paragraph{Writing optimization problems.}
In this class, optimization problems can be introduced in the sentence as
nouns. For example, write as follows:
\begin{quote}
Consider the problem
    \begin{equation}
    \begin{array}{ll}
    \mbox{minimize}   & (1/2)\|Ax-b\|_2^2 + \lambda \|x\|_1 \\
    \mbox{subject to} & 0 \preceq x \preceq \ones \\
    & \|x\|_2 \leq 1,
    \end{array}
    \label{e-opt-prob}
    \end{equation}
where $x \in \reals^n$ is the optimization variable, and $A \in \reals^{m
\times n}$, $b \in \reals^m$, and $\lambda > 0$ are problem data.
\end{quote}
It is important to state which symbols refer to variables and which to problem data.

Be sure to carefully distinguish an optimization problem, 
an algorithm for solving it, and its optimal value.
You \emph{solve} a problem (possibly, using a particular algorithm);
an optimization problem \emph{has} an optimal value;
and an optimization problem \emph{is} convex, or infeasible.
It does not make sense to talk about whether or not a problem converges.

\label{s-content}

\end{document}

### 2015-01-05 code/build/lib/completethat/__init__.py

@@ -1,0 +1,29 @@
+""" CompleteThat is a python package that solves the low rank matrix completion
+problem. Given a low rank matrix with partial entries the package solves an
+optimization problem to estimate the missing entries.
+
+Mathematically, the package solves a relaxation (using the nuclear norm or the
+Frobenius norm of the objective matrix) of the following problem:
+    minimize_{X} ||X||
+    st. X(i,j) = M(i,j) \forall (i,j)\in \Omega,
+
+Where, M represents the data matrix and \Omega represents the set of p observed entries of M
+
+    Usage:
+    # MatrixCompletion
+    >>> from completethat import MatrixCompletion
+    >>> problem = MatrixCompletion(M) 
+    >>> problem.complete_it(algo_name) 
+    >>> X = problem.get_matrix() 
+    >>> out_info = problem.get_out() #Extra info (iterations, ect)
+
+    # MatrixCompletionBD
+    >>> from completethat import MatrixCompletionBD 
+    >>> temp=MatrixCompletionBD('input_data.txt')
+    >>> temp.train_sgd(dimension=6,init_step_size=.01,min_step=.000001, reltol=.001,rand_init_scale=10,   maxiter=1000,batch_size_sgd=50000,shuffle=True):
+    >>> temp.validate_sgd('test_data.txt')
+    >>> temp.save_model()
+
+"""
+from matrix_completion import MatrixCompletion
+from matrix_completion import MatrixCompletionBD

### 2015-01-05 code/build/lib/completethat/matrix_completion.py

@@ -1,0 +1,493 @@
+import numpy as np
+from scipy.sparse import csc_matrix
+from scipy.sparse import linalg as linalg_s
+import os, random, time
+from scipy import linalg
+
+class MatrixCompletion:
+    """ A general class to represent a matrix completion problem
+
+    Data members 
+    ==================== 
+    M:= data matrix (numpy array).
+    X:= optimized data matrix (numpy array)
+    out_info:= output information for the optimization (list) 
+
+
+    Class methods
+    ====================
+    complete_it():= method to complete the matrix
+    get_optimized_matrix():= method to get the solution to the problem
+    get_matrix():= method to get the original matrix
+    get_out():= method to get extra information on the optimization (iter
+    number, convergence, objective function)
+
+    """
+    def __init__(self, X,*args, **kwargs):
+        """ Constructor for the problem instance
+
+            Inputs:
+             1) X: known data matrix. Numpy array with np.nan on the unknow entries. 
+                example: 
+                    X = np.random.randn(5, 5)
+                    X[1][3] = np.nan
+                    X[0][0] = np.nan
+                    X[4][4] = np.nan
+
+        """
+        # Initialization of the members
+        self._M = X
+        self._X = np.array(X, copy = True) #Initialize with ini data matrix
+        self._out_info = []
+
+    def get_optimized_matrix(self):
+        """ Getter function to return the optimized matrix X 
+
+            Ouput:
+             1) Optimized matrix
+        """
+        return self._X
+
+    def get_matrix(self):
+        """ Getter function that returns the original matrix M
+
+            Output:
+            1) Original matrix M
+        """
+        return self._M
+
+    def get_out(self):
+        """ Getter function to return the output information 
+            of the optimization
+
+            Output:
+             1) List of length 2: number of iterations and relative residual
+
+        """
+        return self._out_info
+
+
+    def _ASD(self, M, r = None, reltol=1e-5, maxiter=5000):
+        """
+        Alternating Steepest Descent (ASD)
+        Taken from Low rank matrix completion by alternating steepest descent methods
[…]

### 2015-01-05 code/build/lib/completethat/__init__.py
futher change in structure of files

@@ -1,29 +1,0 @@
-""" CompleteThat is a python package that solves the low rank matrix completion
-problem. Given a low rank matrix with partial entries the package solves an
-optimization problem to estimate the missing entries.
-
-Mathematically, the package solves a relaxation (using the nuclear norm or the
-Frobenius norm of the objective matrix) of the following problem:
-    minimize_{X} ||X||
-    st. X(i,j) = M(i,j) \forall (i,j)\in \Omega,
-
-Where, M represents the data matrix and \Omega represents the set of p observed entries of M
-
-    Usage:
-    # MatrixCompletion
-    >>> from completethat import MatrixCompletion
-    >>> problem = MatrixCompletion(M) 
-    >>> problem.complete_it(algo_name) 
-    >>> X = problem.get_matrix() 
-    >>> out_info = problem.get_out() #Extra info (iterations, ect)
-
-    # MatrixCompletionBD
-    >>> from completethat import MatrixCompletionBD 
-    >>> temp=MatrixCompletionBD('input_data.txt')
-    >>> temp.train_sgd(dimension=6,init_step_size=.01,min_step=.000001, reltol=.001,rand_init_scale=10,   maxiter=1000,batch_size_sgd=50000,shuffle=True):
-    >>> temp.validate_sgd('test_data.txt')
-    >>> temp.save_model()
-
-"""
-from matrix_completion import MatrixCompletion
-from matrix_completion import MatrixCompletionBD

### 2015-01-05 code/build/lib/completethat/matrix_completion.py

@@ -1,493 +1,0 @@
-import numpy as np
-from scipy.sparse import csc_matrix
-from scipy.sparse import linalg as linalg_s
-import os, random, time
-from scipy import linalg
-
-class MatrixCompletion:
-    """ A general class to represent a matrix completion problem
-
-    Data members 
-    ==================== 
-    M:= data matrix (numpy array).
-    X:= optimized data matrix (numpy array)
-    out_info:= output information for the optimization (list) 
-
-
-    Class methods
-    ====================
-    complete_it():= method to complete the matrix
-    get_optimized_matrix():= method to get the solution to the problem
-    get_matrix():= method to get the original matrix
-    get_out():= method to get extra information on the optimization (iter
-    number, convergence, objective function)
-
-    """
-    def __init__(self, X,*args, **kwargs):
-        """ Constructor for the problem instance
-
-            Inputs:
-             1) X: known data matrix. Numpy array with np.nan on the unknow entries. 
-                example: 
-                    X = np.random.randn(5, 5)
-                    X[1][3] = np.nan
-                    X[0][0] = np.nan
-                    X[4][4] = np.nan
-
-        """
-        # Initialization of the members
-        self._M = X
-        self._X = np.array(X, copy = True) #Initialize with ini data matrix
-        self._out_info = []
-
-    def get_optimized_matrix(self):
-        """ Getter function to return the optimized matrix X 
-
-            Ouput:
-             1) Optimized matrix
-        """
-        return self._X
-
-    def get_matrix(self):
-        """ Getter function that returns the original matrix M
-
-            Output:
-            1) Original matrix M
-        """
-        return self._M
-
-    def get_out(self):
-        """ Getter function to return the output information 
-            of the optimization
-
-            Output:
-             1) List of length 2: number of iterations and relative residual
-
-        """
-        return self._out_info
-
-
-    def _ASD(self, M, r = None, reltol=1e-5, maxiter=5000):
-        """
-        Alternating Steepest Descent (ASD)
-        Taken from Low rank matrix completion by alternating steepest descent methods
[…]

### 2015-01-05 document/article.tex
add matrix completion problem

@@
 \section{Matrix Completion Problem}
+The matrix completion problem can be formally stated as follows.
+We are interested in recovering a matrix $M \in \reals^{n_1 \times n_2}$ but only get to observe a number $m \ll n_1n_2$ of its entries.
+Thus, we want to find a solution to the following optimization problem
 
-\begin{quote}
-Consider the problem
 \begin{equation}
     \begin{array}{ll}
-    \mbox{minimize}   & (1/2)\|Ax-b\|_2^2 + \lambda \|x\|_1 \\
-    \mbox{subject to} & 0 \preceq x \preceq \ones \\
-    & \|x\|_2 \leq 1,
+    \mbox{minimize}   &  \rank(X)\\
+    \mbox{subject to} & X_{ij} = M_{ij} \qquad (i,j)\in \Omega
     \end{array}
@@
     \end{equation}
-where $x \in \reals^n$ is the optimization variable, and $A \in \reals^{m
-\times n}$, $b \in \reals^m$, and $\lambda > 0$ are problem data.
-\end{quote}
+where $X\in \reals^{n_1 \times n_2}$ is the decision variable and $\rank(X)$ is equal to the rank of the matrix $X$.
 
@@
 
-\caption{ASD method applied three different Columbia University images}
+\caption{ASD method applied to three different Columbia University images}
 \end{figure}
@@
 \newpage
-
-\paragraph{Punctuation in equations.}
-An equation is part of a sentence, and you may need to include a comma or a period
-at the end of an equation as a result, whether or not you are using inline or
-display math style. For example:
-\begin{quote}
-    We next discuss how to solve the problem
-    \[
-        \begin{array}{ll}
-            \mbox{minimize} & (1/2)\|Ax - b\|_2^2,
-        \end{array}
-    \]
-    where $x \in \reals^n$ is the optimization variable.
-\end{quote}
-
-\paragraph{Don't start a sentence with a symbol.}
-This hurts readability:
-\begin{quote}
-Bad: $f$ is smooth.\\
-Good: The function $f$ is smooth.
-
-Bad: $x^n - a$ has $n$ distinct zeros. \\
-Good: The polynomial $x^n - a$ has $n$ distinct zeros.
-\end{quote}
-Similarly, don't start a sentence with a reference.
-
-\paragraph{Use words to separate symbols in different formulas.}
-If it might confuse the reader visually or in the actual
-meaning of the sentence, use words to break apart formulas:
-\begin{quote}
-Bad: The sequences $x_1, x_2, \dots$, $y_1, y_2, \dots$ are Cauchy. \\
-OK: The sequences $x_1, x_2, \dots,$ and $y_1, y_2, \dots,$ are Cauchy. \\
-Good: The sequences $(x_i)$ and $(y_i)$ are Cauchy.
-
-OK: The image of $S$ under $f$, $f(S) = \{ x \mid x \in S \}$, is convex. \\
-Good: The image of $S$ under $f$, given by $f(S) = \{ x \mid x \in S \}$, is convex.
-\end{quote}
-Do not insert superfluous words if the meaning is clear.
-\begin{quote}
-Good: Consider the function $f + g + h$, where $f : \reals^n \to
-\reals$, $g : \reals^m \to \reals$, and $h : \reals^p \to \symm^n$ are closed
-proper convex.
-\end{quote}
-
-\paragraph{References.}
-For internal references, use the \texttt{label} and \texttt{ref} commands.
-\emph{Never} number or refer to an entity using a specific number,
-as in ``Table~3.''
-Refer to ``Table~\ref{t-ex1}'', ``\eqref{e-opt-prob}'' (an equation), and so on.
-Only number equations that are important and should be emphasized or that you
-specifically refer to elsewhere. Do not simply number all equations, and do not
-number them randomly. A numbered equation draws the reader's attention, so it
-should be used relatively sparingly. Use \verb+\S+ for section references, as in
-\S\ref{s-content}.
-
-For external references, you must use Bib\TeX. Use references like nouns or 
-footnotes:
-\begin{quote}
-Good: One useful reference is \cite{BoV:04}. \\
-Good: Interested readers may refer to Boyd and Vandenberghe \cite{BoV:04}.
-\end{quote}
-
-Your Bib\TeX\ source must be correct.  Unfortunately, many systems that export
-Bib\TeX\ entries export very poor ones.  Bib\TeX\ ignores the \texttt{*.bib}
-file's capitalization for articles (but not books). To force capitalization,
-you must wrap the letter with curly braces.  The hyphen (-), en dash (--), and
-em dash (---) are distinct punctuation marks that serve different purposes.
-(And the minus sign is different from all three.) When specifying a page range
-in references, use the en dash. For an example of escaping and using the en
-dash, look at our Bib\TeX\ entry for \cite{Tref:2008}.
 
@@
 
-\paragraph{Use of notation and jargon.}
-The correct and appropriate use of notation and jargon takes time to master.
-The following should give you a sense of what to think about:
-
-\begin{itemize}
-\item Don't use the same notation for two different things. Conversely, be consistent
-about notation for the same thing mentioned in two places: don't
-say ``$A_j$ for $1 \leq j \leq n$'' in one place and ``$A_i$ for $i=1,\ldots,n$'' 
-in another, or use two different indices in summations over the same ranges. 
-Formally, the $i$ and $j$ are dummy variables with only that phrase as scope,
-so these are technically correct, but it will confuse the reader.
-
-\item It can be useful to choose names for indices so, for example, $i$
-always varies from 1 to $m$ and $j$ always varies from 1 to $n$ (when
-referencing something like the rows and columns of an $m \times n$ matrix).
-
-\item Define all symbols before or near to where you use them.  A good rule is that
-when you first use something, say, $X$, you should either define it immediately, or 
-within the paragraph.   There are a few cases where it is acceptable to define it 
-later, but you must say this explicitly, as in ``$X$ is the matrix of activation
-levels, which we define below.''
-
-\item A symbol like $f$ refers to a function, while $f(x)$ refers to a
-function evaluated at a given point. Avoid sloppy writing like ``The function
-$f(x)$ is convex.''
-So-called `anonymous' functions defined inline are an exception to this rule,
-as in ``the function $x^2 \cos x$ is a counterexample,'' though this should
-be used only for simple functions.
-
-\item Use typographic naming conventions like capital letters for sets, Greek
-letters for dual variables, \emph{etc}.
-
-\item Try to use mnemonic notation, so $x_c$ for a center point,
-$c$ for a cost vector, $S$ for a generic set, $C$ for a convex set, $B$
-for a ball, and so on. Note that there are conventions about
-the use of different letters in mathematics that you should not ignore:
-$i$, $j$, and $k$ are usually used for indices, $x$ through $z$ for
-variables, $a$ through $d$ for constants, \emph{etc}.
-
-\item Don't use symbols like $\forall$, $\exists$, and $\Rightarrow$; use the
-corresponding words. These symbols are usually appropriate only in
-formal logic.
-
-\item Don't overuse subscripts. If you do not need to assign a subscript
-to something, don't. For example, if you begin by defining a set as
-$X = \{x_1, \dots, x_n\}$, then it is going to get annoying to refer to subsets
-of $X$, since you will need double subscripts. If possible, it would be better
-to not name the elements of $X$ and only refer to them when necessary, just
-as, say, $x$ and $y$.
-
-Beware especially of any topic in graph theory when it comes to this rule;
-there are often \emph{many} ways to refer to relevant objects, and finding
-reasonable notation can often get one half the way to getting a handle on the
-problem itself.
-
-\item Don't assign symbols to concepts that you never refer to, or can easily
-refer to without:
-\begin{quote}
-Bad: Let $X$ be a compact subset of a space $Y$. If $f$ is a continuous
-real-valued function over $X$, it has a minimum over $X$. \\
-Good: A continuous real-valued function has a minimum over a compact set.
-\end{quote}
-Similarly, do not say ``The solution $x^\star$ is unique'' if you never need
-to refer to $x^\star$ again; simply say that the solution is unique.
-When you say ``The solution $x^\star$ is unique'' you are both stating a fact
-\emph{and} entering the symbol $x^\star$ into the paper's symbol table and
-the reader's working memory.
-\end{itemize}
-
-\paragraph*{Sections.}
-Use structures like \texttt{section}, \texttt{subsection},
-\texttt{subsubsection}, and \texttt{paragraph} to organize the exposition;
-never insert manual line breaks or page breaks for this purpose. In particular,
-do not use line breaks to separate different topics. Do not use symbols in
-article titles and ideally not even in section headers.
-
-\paragraph{Use capitalization consistently.}
-For section headers, capitalize the first letter, proper nouns, and lowercase
-everything else, as in this document.  For references, write Figure 1, Algorithm 3,
-Theorem 4, and so on.
-
 \paragraph{Use the right commands.}
@@
 
-\paragraph{Writing optimization problems.}
-In this class, optimization problems can be introduced in the sentence as
-nouns. For example, write as follows:
-\begin{quote}
-Consider the problem
-    \begin{equation}
-    \begin{array}{ll}
-    \mbox{minimize}   & (1/2)\|Ax-b\|_2^2 + \lambda \|x\|_1 \\
-    \mbox{subject to} & 0 \preceq x \preceq \ones \\
-    & \|x\|_2 \leq 1,
-    \end{array}
-    \label{e-opt-prob}
-    \end{equation}
-where $x \in \reals^n$ is the optimization variable, and $A \in \reals^{m
-\times n}$, $b \in \reals^m$, and $\lambda > 0$ are problem data.
-\end{quote}
-It is important to state which symbols refer to variables and which to problem data.
-
-Be sure to carefully distinguish an optimization problem, 
-an algorithm for solving it, and its optimal value.
-You \emph{solve} a problem (possibly, using a particular algorithm);
-an optimization problem \emph{has} an optimal value;
-and an optimization problem \emph{is} convex, or infeasible.
-It does not make sense to talk about whether or not a problem converges.
-
 \label{s-content}

### 2015-01-05 document/article.tex
update document

@@
     \end{array}
-    \label{eq:original_problem}
+    \label{eq:original_problem1}
 \end{equation}
 where $X\in \reals^{n_1 \times n_2}$ is the decision variable and $\rank(X)$ is equal to the rank of the matrix $X$.
+The problem (\ref{eq:original_problem1}) seeks the simplest matrix fitting the observed data.
+Of course, if there were only one low-rank object fitting the data, this would recover $M$.
+In order to fix the notation correctly define the projector $P_{\Omega}:\reals^{n_1\times n_2} \rightarrow \reals^{n_1\times n_2} $ as
+
+\begin{equation*}
+P_{\Omega}(A) = \left \{ 
+	\begin{array}{lr}
+		A_{ij} & : (i,j) \in \Omega\\
+		0 & : \mathrm{otherwise}
+	\end{array}
+	\right.
+\end{equation*}\\
+Using the projector, Problem (\ref{eq:original_problem1}) can be rewritten as
+\begin{equation}
+    \begin{array}{ll}
+    \mbox{minimize}   &  \rank(X)\\
+    \mbox{subject to} & P_{\Omega}(X) = P_{\Omega}(M)
+    \end{array}
+    \label{eq:original_problem2}
+\end{equation}
+
+Unfortunately, this optimization problem has been shown to be NP-hard and all know algorithms which provide exact solutions require time doubly exponential in the dimension of the matrix both in theory and in practice~\cite{Candes:2009}.
+Also, the $\rank(X)$ makes the problem non-convex.
+
+Several approximations to the problem exists. One, the ``tightest convex relaxation of $\rank(X)$''~\cite{Fazel:2002} is the following problem
+
+\begin{equation}
+    \begin{array}{ll}
+    \mbox{minimize}   &  \|X\|_* :=\sum_{i=1}^r \sigma_i(X) \\
+    \mbox{subject to} & P_{\Omega}(X) = P_{\Omega}(M)
+    \end{array}
+    \label{eq:nucl_problem}
+\end{equation}
+
+where $\sigma_i(X)$ denotes the $i$th largest singular value of $X$ and $\|X\|_*$ is the called the nuclear norm.
+The main point of this relaxation is that the nuclear norm is a convex function and thus can be optimized efficiently via semidefinite programming or by iterative soft thresholding algorithms.
+
+Alternative to nuclear norm minimization there have been many algorithms which are designed to target the following optimization problem
+\begin{equation}
+    \begin{array}{ll}
+    \mbox{minimize}_{X,Y}   & \frac{1}{2} \|P_{\Omega}(XY) - P_{\Omega}(M)\|^2_F \\
+    \end{array}
+    \label{eq:frob_problem}
+\end{equation}
+where $X=YZ$ is the completed matrix and $Y\in \reals^{n_1\times r}, Z\in \reals^{r\times n_2}$ and $r$ represents the (hopefully small) rank of the matrix $X$.
 
@@
 \subsection*{CVX}
-Problem (\ref{eq:original_problem}) can be easily solved directly using CVX, a package for specifying and solving convex programs~\cite{cvx}, ~\cite{gb08}, with the following code:
+Problem (\ref{eq:nucl_problem}) can be easily solved directly using CVX, a package for specifying and solving convex programs~\cite{cvx}, ~\cite{gb08}, with the following code:
 
@@
 
+\clearpage
 \subsection*{Columbia University images}
 
-\newpage
 \begin{figure}[H]
+\resizebox{\textwidth}{!}{%
 \makebox[\textwidth][c]
@@
 \end{tabular}
-
+}
 }
@@
 \end{figure}
-
-\newpage
 
@@
 
-\newpage
-
-\paragraph{Do not italicize English in math mode.}
-Mathematical symbols should be typeset in math mode:
-write $Ax=b$, not Ax=b. This said, subscripts or
-superscripts that derive from English (or any human language) should not be
-italicized. For example, write $f_\mathrm{best}$, not $f_{best}$. The exception
-is subscripts based on a single letter: refer
-to a point that is the center of some set as $x_c$, not $x_{\mathrm{c}}$.
-Similarly, use commands for special functions: use $\sin(x)$,
-$\log(x)$, and $\exp(x)$, not $sin(x)$, $log(x)$, or $exp(x)$.
-
-A really heinous example would be the following:
-\begin{quote}
-Consider the problem
-\[
-\begin{array}{ll}
-minimize & f(Ax - b) \\
-\end{array}
-\]
-where x is the optimization variable and A and b are problem data.
-\end{quote}
-
-\paragraph{Spacing.}
-A blank line ends a paragraph. You shouldn't leave a blank line between an
-equation and the following text unless you intend the equation to end the
-paragraph. Write:
-\begin{quote}
-\begin{verbatim}
-The image of $S$ under $f$,
-\[
-    f(S) = \{ f(x) \mid x \in S \},
-\]
-is convex.
-\end{verbatim}
-\end{quote}
-Inserting extra blank lines before \verb+\[+ or after \verb+\]+ will result in
-bad typesetting. However, the following is fine, since a new paragraph is called for:
-\begin{quote}
-\begin{verbatim}
-The image of $S$ under $f$ is defined as
-\[
-    f(S) = \{ f(x) \mid x \in S \}.
-\]
-
-We now turn to a different topic.
-\end{verbatim}
-\end{quote}
-Using a tab before \verb+f(S)+ inside the equation environment is
-optional, but helps readability in the source.  Similar rules should be
-followed for other environments like \texttt{quote}; see the source of this
-document.
-
-\paragraph{Use the right commands.}
-There are certain special commands in \LaTeX\ for notation that you otherwise
-might attempt to write in an ad-hoc manner. If you do the latter, the typesetting
-will be inferior. A few common examples:
-\begin{itemize}
-\item Norms are written $\|x\|$, not $||x||$.
-
-\item For set-builder notation, use
-$\{ x \in \reals \mid x \geq 0 \}$, not $\{x \in \reals | x \geq 0 \}$.
-
-\item Make sure to use \verb+\left+ and \verb+\right+ when wrapping
-taller expressions with parentheses, curly braces, brackets, and so on.
-\begin{quote}
-Bad: Let
-\[
-    x = \argmin_u (f(u) + \frac{1}{2}\|u - z\|_2^2).
-\]
-Good: Let
-\[
-    x = \argmin_u \left(f(u) + \frac{1}{2}\|u - z\|_2^2\right).
-\]
-\end{quote}
-
-\item Use \verb+\ldots+ (lower dots) when the dots are surrounded by commas and
-\verb+\cdots+ (center dots) when surrounded by other objects that have full
-height, as in $x_1, x_2, \ldots, x_n$ and $x_1 + x_2 + \cdots + x_n$.
-\end{itemize}
-Similarly, make sure to use our conventions for standard mathematical objects
-for this class; for example, use $\reals$, not $\mathbb{R}$, to denote the set
-of real numbers. Use the \verb+\argmin+ and \verb+\argmax+ commands (as shown
-above); do not write `arg min', since `argmin' is a single mathematical
-operator.  These and other definitions are provided in \verb+defs.tex+.
-
-\label{s-content}
-
 \end{document}

### 2015-01-05 document/article.tex
update document

@@
 We briefly discuss the theory behind matrix factorizing, the models and approaches we choose to use and the algorithms for the optimization.
-Later, we walk through two case studies illustrating the usefulness and applicability in the context of image process and a recommendation system for Yahoo music and movie data.
+Later, we walk through two case studies illustrating the usefulness and applicability in the context of image processing and a recommendation system for Yahoo music and movie data.
 
@@
 where $\sigma_i(X)$ denotes the $i$th largest singular value of $X$ and $\|X\|_*$ is the called the nuclear norm.
-The main point of this relaxation is that the nuclear norm is a convex function and thus can be optimized efficiently via semidefinite programming or by iterative soft thresholding algorithms.
+The main point of this relaxation is that the nuclear norm is a convex function and thus can be optimized efficiently via semidefinite programming or by iterative soft thresholding algorithms~\cite{cai:2010}~\cite{goldfarb:2011}.
 
@@
     \begin{array}{ll}
-    \mbox{minimize}_{X,Y}   & \frac{1}{2} \|P_{\Omega}(XY) - P_{\Omega}(M)\|^2_F \\
+    \mbox{minimize}_{Y,Z}   & \frac{1}{2} \|P_{\Omega}(YZ) - P_{\Omega}(M)\|^2_F \\
     \end{array}
@@
 \clearpage
-\subsection*{Columbia University images}
+\subsection*{Columbia University photographs}
+In this section, we demonstrate the applicability of the package to image processing problems.
+By randomly erasing a predefined proportion of the pixels of an image the problem of completing the image can be cast as a matrix completion problem.
+For illustration we used three  512 x 512 grayscale photographs taken by one of the authors of the Columbia University campus on December, 2014.
+The reconstructed images can be seen on Figure~\ref{fig:columbia_pics}, where we erased 65\% of the pixels of each image.
+
+In addition, to illustrate the ease of use of the package we also present the script used to solve all the problems in what follows:
+
+\lstinputlisting[numbers=none]{columbia_images.py}
 
@@
 
-\caption{ASD method applied to three different Columbia University images}
+\caption{ASD method applied three different Columbia University photographs using random sampling}
+\label{fig:columbia_pics}
 \end{figure}

### 2015-01-05 document/article.tex
update document
updated document: algorithms, photographs section and overall structure

@@
 \usepackage{booktabs}
+\usepackage{algpseudocode, algorithm}
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
+\renewcommand{\algorithmicuntil}{\textbf{Until:}}
+\renewcommand{\algorithmicrepeat}{\textbf{Repeat:}}
 
@@
 
-\title{\texttt{CompleteThat}: A python package for  low-rank matrix completion}
+\title{\texttt{CompleteThat}: A python package for  low-rank matrix completion\\EEOR E4650 Course Project
+}
 \author{Joshua Edgerton \and Esteban Fajardo}
@@
 where $X=YZ$ is the completed matrix and $Y\in \reals^{n_1\times r}, Z\in \reals^{r\times n_2}$ and $r$ represents the (hopefully small) rank of the matrix $X$.
+The main point of this relaxation is the use of an alternating minimization scheme or Gauss-Seidel or 2-block coordinate descent method, where first $Y$ is fixed and then the problem is reduced to a convex, standard least squares problem on $Z$.
+Later, $Z$ is fixed and then the problem is reduced to a convex, standard least squares problem on $Y$.
 
@@
 However, the computation is very slow and for any matrix greater than 100x100 the computational time is too high.
-Since this is clearly unsatisfactory for practical purposes, specific algorithms were developed and used on the package.
+Since this is clearly unsatisfactory for practical purposes, specific algorithms were implemented on the package.
 
@@
 \subsubsection*{Low rank matrix completion by alternating steepest descent methods}
+The  alternating steepest descent methods implemented on \texttt{CompleteThat} solve the Problem (\ref{eq:frob_problem}) above.
+Let the objective function of the optimization problem $f: \reals^{n_1\times r}\times \reals^{r\times n_2} \rightarrow \reals$ be defined as
+\begin{equation}
+	f(Y,Z):=\frac{1}{2} \|P_{\Omega}(YZ) - P_{\Omega}(M)\|^2_F.
+	\label{eq:ASD_prob}
+\end{equation}
+
+To solve efficiently the above problem, Tanner and Wei~\cite{Tanner:2014} use a single step of simple line-search along the gradient descent directions.
+Let us write $f(Y, Z)$ as $f_Z(Y)$ when $Z$ is held constant and $f_Y(Z)$ when $Y$ is held constant and let $t_y, t_z$ be the steepest descent step sizes for descent directions.
+
+The Alternating Steepest Descent Method (ASD) applies steepest gradient descent to $f(Y, Z)$ in (\ref{eq:ASD_prob}) alternatively with respect to $Y$ and $Z$.
+It can be shown that the directions of gradient ascent and the steepest descent step sizes are~\cite{Tanner:2014} :
+\[
+\nabla f_{Z}(Y) = -(P_{\Omega}(M) - P_{\Omega}(Y Z))Z^T\
+\]
+\[
+\nabla f_{Y}(Z) = -Y^T(P_{\Omega}(M)-P_{\Omega}(YZ)).
+\]
+\[
+t_{y} = \frac{\|  \nabla f_{Z}(Y) \|^2_F}{\| P_{\Omega}(\nabla f_{Z} (Y)Z\|^2_F}
+\]
+\[
+ t_{z} = \frac{\|  \nabla f_{Y}(Z) \|^2_F}{\| P_{\Omega}(Y \nabla f_Y (Z)) \|^2_F}
+\]
+
+\begin{algorithm}[h]
+\caption{Alternating Steepest Descent (ASD)~\cite{Tanner:2014}}
+\begin{algorithmic}
+\Require $r \in \reals, P_{\Omega}(M), Y_0\in \reals^{n_1\times r}, Z_0\in \reals^{r\times n_2}$
+\Repeat
+	\State $\nabla f_{Z_i}(Y_i) = -(P_{\Omega}(M) - P_{\Omega}(Y_i Z_i))Z_i^T, t_{y_i} = \frac{\|  \nabla f_{Z_i}(Y_i) \|^2_F}{\| P_{\Omega}(\nabla f_{Z_i} (Y_i)Z_i \|^2_F}$
+	\State $Y_{i+1} = Y_i - t_{y_i} f_{Z_i}(X_i)$
+	\State $\nabla f_{Y_{i+1}}(Z_i) = -Y^T_{i+1}(P_{\Omega}(M)-P_{\Omega}(Y_{i+1}Z_i)), t_{z_i} = \frac{\|  \nabla f_{Y_{i+1}}(Z_i) \|^2_F}{\| P_{\Omega}(Y_{i+1} \nabla f_Y{i+1} (Z_i)) \|^2_F}$
+	\State $Z_{i+1} = Z_i - t_{z_i}\nabla f_{Y_{i+1}}(Z_i)$
+	\State $i = i +1$
+\Until {termination criteria is not reached}
+\Ensure $X:= Y_i Z_i \in \reals^{n_1\times n_2}$
+\end{algorithmic}
+\end{algorithm}
+
+The Scaled Alternating Steepest Descent Method (sASD) is an accelerated version of ASD and uses a scaled gradient descent direction with exact line-search.
+Its main motivation for the scaled factors is the explicit expression for the Newton directions for the problem (\ref{eq:ASD_prob}) if all the entries in the matrix $M$ were know:
+\[
+(M - YZ)Z^T(ZZ^T)^{-1}
+\]
+\[
+(X^TX)^{-1}X^T(M-YZ)
+\]
+which are the gradient descent directions scaled by $(ZZ^T)^{-1}$ and $(X^TX)^{-1}$.
+
+\begin{algorithm}[h]
+\caption{Scaled Alternating Steepest Descent (sASD)~\cite{Tanner:2014}}
+\begin{algorithmic}
+\Require $r \in \reals, P_{\Omega}(M), Y_0\in \reals^{n_1\times r}, Z_0\in \reals^{r\times n_2}$
+\Repeat
+	\State $\nabla f_{Z_i}(Y_i) = -(P_{\Omega}(M) - P_{\Omega}(Y_i Z_i))Z_i^T$
+	\State $d_{y_i} = -\nabla f_{Z_i} (Y_i)(Z_iZ_i^T)^{-1}, t_{y_i} =\frac{ -\nabla f_{Z_i}(Y_i)\cdot d_{y_i}}{\| P_{\Omega} (d_{y_i}Z_i)\|_F^2}$
+	\State $Y_{i+1} = Y_i + t_{y_i}d_{y_i}$
+	\State $\nabla f_{Y_{i+1}} (Z_i) = - Y_{i+1}^T(P_{\Omega}(M) - P_{\Omega}(Y_{i+1}Z_i))$
+	\State $d_{z_i} = (Y_{i+1}^TY_{i+1}^{-1} \nabla f_{Y_{i+1}}(Z_i), t_{z_i} = \frac{-\nabla f_{Y_{i+1}}(Z_i)\cdot d_{z_i}}{\| P_{\Omega}(Y_{i+1}d_{z_i})\|_F^2}$
+	\State $Z_{i+1} = Z_i + t_{z_i} d_{z_i}$
+	\State $i = i + 1$
+\Until {termination criteria is not reached}
+\Ensure $X:= Y_i Z_i \in \reals^{n_1\times n_2}$
+\end{algorithmic}
+\end{algorithm}
+
 \subsubsection*{Low rank matrix completion by stochastic gradient descent}
+The formal problem as defined in the model section is stated in the language of linear algebra and the typical optimizations using gradient methods require access to all of the data at once for updating the parameters, since its dependent on calculating inverse of matrices or solving systems of linear equations.
+However, given the scale of data that is common in today's digital world, the need for more efficient and memory friendly algorithms arises.
+With stochastic gradient descent, a variation of the standard batch gradient optimization, instead of calculating the gradient for the user feature vector using all the item feature vectors that the user has rated as would be done in usually, we approximate the user's gradient using the current item vector for the current rating and vice versa for updates of the item vectors.
+Thus we are essentially exchanging the luxury of using less memory and computational resources for more time and iterations spent with stochastic gradient descent.\footnote{We note that our formulation is not convex but, due to shuffling of the data and the way the gradients are estimated, the algorithm will converge to at least a local minimum.}
+
+\begin{algorithm}[H]
+\caption{Stochastic Gradient Descent (SGD)}
+\begin{algorithmic}
+\Require
+\State Initialize Parameters
+\Repeat
+	\State Read next record of file
+	\State Lookup user and item keys in dictionary
+	\State Update user and item feature vectors
+	\State Update the total sum of errors variable
+
+\Until {termination criteria is not reached}
+\State Shuffle Data Set
+\Ensure
+\end{algorithmic}
+\end{algorithm}
+
+\paragraph{Notes}
+\begin{itemize}
+
+	\item Step Size: The algorithms performance is highly dependent on the step size (alpha) parameter of the update equations.
+Too small of a alpha will result in convergence taking a very long time and lower the chance of finding the global objective.
+Too large of a alpha will result in convergence not happening as the the algorithm continually 'overshoots' the optimal value.
+
+	\item Shuffle: When dealing with large data we need to be clever with how we shuffle our data.
+Shuffling is no trivial task out of memory.
+We explored several out of memory methods but ultimately decided with using a pseudo-shuffle, where we read in the file in large chunks that fit in memory, shuffle the lines using built in python functions, write the records back to a text file, and continue until the all of the file has been pseudo-shuffled.
+Hopefully by choosing a large enough batch size for reading in, our file is shuffle well enough.
+We also explored with a more robust method of shuffling the shuffled batches, but the added time complexity of appending large files together did not prove worthwhile given almost negligible returns in convergence.
+
+\end{itemize}
 
@@
 \subsubsection*{MatrixCompletion}
-Given a \texttt{numpy} matrix $M$ with \texttt{numpy.nan} on the missing entries, the matrix completion problem can be solved (using ASD) as:
+Given a \texttt{numpy} matrix $M$ with \texttt{numpy.nan} on the missing entries, the matrix completion problem can be solved (using ASD, for example) as:
 \begin{verbatim}
@@
 
-We look to evaluate the performance of our algorithms on these two datasets, comparing the various algorithms implemented versus a naive benchmark estimate where we used the mean of the training set as our estimator for all records in the test set.
+We look to evaluate the performance of our SGD algorithm on these two datasets comparing it versus a naive benchmark estimate, where we used the mean of the training set as our estimator for all records in the test set.
 Then we compared how well the algorithms performed versus the benchmark.
@@
 A quick run of Mahout yields 1.17 and CompleteThat's sgd model yielded a 1.19.
-So our model yielded a pretty similar RMSE as the MAHOUT package.
+So our model yielded a very similar RMSE as the MAHOUT package.
 Note that for both of these instances we used a factorization of rank equal to seven and the Complete that sgd algorithm ran slightly faster than Mahout's (.537 minutes vs 1.4 minutes).
@@
 In this section, we demonstrate the applicability of the package to image processing problems.
-By randomly erasing a predefined proportion of the pixels of an image the problem of completing the image can be cast as a matrix completion problem.
+Grayscale pictures can be transform into a rectangular matrix where each element of the matrix determines the intensity of the corresponding pixel.
+For convenience, most of the current digital files use integer numbers between 0 and 255.
+By randomly erasing, or setting to zero, a predefined proportion of the pixels of an image the problem of completing the image can be cast as a matrix completion problem, where the missing entries are the those with zero on it.
+
 For illustration we used three  512 x 512 grayscale photographs taken by one of the authors of the Columbia University campus on December, 2014.
-The reconstructed images can be seen on Figure~\ref{fig:columbia_pics}, where we erased 65\% of the pixels of each image.
-
-In addition, to illustrate the ease of use of the package we also present the script used to solve all the problems in what follows:
+The original, erased and reconstructed images are shown on Figure~\ref{fig:columbia_pics}, where we erased 65\% of the pixels of each image.
+In addition, to illustrate the ease of use of the package, we also present the (very short) script used to recover the photographs:
 
@@
 \section{Source code}
-On the whole, the package directory structure looks like the Figure~\ref{fig:code_scheme}. In the following we present the source code for the package.
+On the whole, the package directory structure looks like Figure~\ref{fig:code_scheme}. In what follows we present the source code of the package.
 \begin{figure}[h!]
@@
 \code{completethat/matrix\_\_init\_\_.py}{../code/completethat/__init__.py}
-\code{completethat/matrix\_\_init\_\_.py}{../code/setup.py}
+\code{setup.py}{../code/setup.py}
+\hrulefill
+\subsection*{README.rst}
+\lstinputlisting{../code/README.rst}
 

### 2015-01-05 document/columbia_images.py

@@ -1,0 +1,30 @@
+import numpy as np
+from scipy import misc
+from completethat import MatrixCompletion
+
+def blur(imgpath, delta = 0.65):
+    """
+        Function to blur the image
+    """
+    photo = scipy.misc.imread(imgpath, flatten=True).astype(float) #Read as grayscale
+    m, n = photo.shape
+    p = round(delta * (m * n - 1)) #Number of blanks
+    A = np.zeros((m * n, 1), dtype=bool)
+    A[0:p] = True
+    ind = np.random.permutation(range(m*n))
+    A = A[ind]
+    A = A.reshape((m,n))
+    return photo, A
+	
+
+if __name__ == '__main__':
+    
+    # Read image and randomly erase 65% of the pixels
+    photo, A = blur('./columbia_1.png')
+    M = np.copy(photo)
+    M[~A] = np.nan
+    problem = MatrixCompletion(M)
+
+    # Solve the problem
+    problem.complete_it('ASD')
+    X = np.copy(problem.get_optimized_matrix())

### 2015-01-05 document/article.tex
updated SGD update document

@@
 	\State $i = i +1$
-\Until {termination criteria is not reached}
+\Until {termination criteria is reached}
 \Ensure $X:= Y_i Z_i \in \reals^{n_1\times n_2}$
@@
 	\State $i = i + 1$
-\Until {termination criteria is not reached}
+\Until {termination criteria is reached}
 \Ensure $X:= Y_i Z_i \in \reals^{n_1\times n_2}$
@@
 However, given the scale of data that is common in today's digital world, the need for more efficient and memory friendly algorithms arises.
-With stochastic gradient descent, a variation of the standard batch gradient optimization, instead of calculating the gradient for the user feature vector using all the item feature vectors that the user has rated as would be done in usually, we approximate the user's gradient using the current item vector for the current rating and vice versa for updates of the item vectors.
+
+Let $U$ be the set of users and $I$ the set of items.
+Let $K\subset U \times I$ be the set of known ratings and $r_{ij}$ the rating that user $i$ gave to item $j, \forall (i,j)\in K$.
+Let $p_i$  be the latent feature vector for user $i$ of length equal to hypothesized rank.
+Let $q_j$ be the latent feature vector for item $j$ of length equal to hypothesized rank.
+The stochastic gradient descent method solves the following optimization problem:
+\begin{equation}
+    \begin{array}{ll}
+    \mbox{minimize}_{r_{ij}\in K}   & f:= (r_{ij} - \hat r_{ij})^2 = e_{ij}^2\\
+    \mbox{subject to} & r_{ij} = p_i^T q_j
+    \end{array}
+    \label{eq:SGD}
+\end{equation}
+
+With stochastic gradient descent, a variation of the standard batch gradient optimization, instead of calculating the gradient for the user feature vector using all the item feature vectors that the user has rated as would be done  usually, we approximate the user's gradient using the current item vector for the current rating and vice versa for updates of the item vectors.
 Thus we are essentially exchanging the luxury of using less memory and computational resources for more time and iterations spent with stochastic gradient descent.\footnote{We note that our formulation is not convex but, due to shuffling of the data and the way the gradients are estimated, the algorithm will converge to at least a local minimum.}
@@
 \begin{algorithmic}
-\Require
-\State Initialize Parameters
+\Require $r_{ij} \in \reals,  \forall (i,j)\in K$
+\State Initialize Parameters: $\forall i\in U$ randomly initialize $p_i\in \reals^r$. $\forall j\in I$ randomly initialize  $q_j\in \reals^r$
 \Repeat
-	\State Read next record of file
-	\State Lookup user and item keys in dictionary
-	\State Update user and item feature vectors
-	\State Update the total sum of errors variable
-
-\Until {termination criteria is not reached}
+	\State Read rating $r_{ij}$
+	\State Compute error $e_{ij} = r_{ij} - p_i^Tq_j$
+	\State Compute gradients: 
+	\State $\qquad \qquad \frac{\partial f}{\partial p_i} = -2e_{ij}q_j$
+	\State $\qquad \qquad \frac{\partial f}{\partial q_j} = -2e_{ij}p_i$
+	\State Update parameters:
+	\State $\qquad \qquad p_i:=p_i + 2\alpha e_{ij} q_j$
+	\State $\qquad \qquad q_j:=q_j + 2\alpha e_{ij} p_i$
+	\State Update MSE: $\text{MSE} := \text{MSE} + e_{ij}$
+\Until {termination criteria is reached}
 \State Shuffle Data Set
-\Ensure
+\Ensure $p_i~~\forall i\in U; q_j ~~\forall j\in I$
 \end{algorithmic}
@@
 \subsection*{Code Example}
-For matrices that fit into memory use the \texttt{MatrixCompletion} module. Otherwise, use \texttt{MatrixCompletionBD} module.
+For matrices that fit into memory use the \texttt{MatrixCompletion} module. Otherwise, use the \texttt{MatrixCompletionBD} module.
 
@@
 For illustration we used three  512 x 512 grayscale photographs taken by one of the authors of the Columbia University campus on December, 2014.
-The original, erased and reconstructed images are shown on Figure~\ref{fig:columbia_pics}, where we erased 65\% of the pixels of each image.
+The original, erased and reconstructed images are shown on Figure~\ref{fig:columbia_pics}, where we erased 35\% of the pixels of each image.
 In addition, to illustrate the ease of use of the package, we also present the (very short) script used to recover the photographs:
@@
 For the memory-based algorithms it is well know that noise in the data introduces overfitting on the estimated matrix.
-Following the approach taken by Mazumder et al.~\cite{mazumder:2010}  where they solve the same objetive function as the problems above (using the Frobenious norm) but introducing the nuclear norm as regularizer to account for overfitting, we would introduce a regularization option into the matrix completion procedure.
+Following the approach taken by Mazumder et al.~\cite{mazumder:2010}  where they solve the same objetive function as the problems above (using the Frobenious norm) but introducing the nuclear norm as regularizer to account for overfitting, we would like to introduce a regularization option into the matrix completion procedure.
 
@@
 
-Finally, given the similarities between matrix completion and robust principal component analysis one further extension to the functionality of the package could be implementing the robust-PCA procedure outlined by Cand\'es~\cite{candes:2011}, et al.
+Finally, given the similarities between matrix completion and robust principal component analysis one further extension to the functionality of the package could be implementing the robust-PCA procedure outlined by Cand\'es et al.~\cite{candes:2011}.
 

### 2015-01-05 document/columbia_images.py

@@ -8,7 +8,7 @@ def blur(imgpath, delta = 0.65):
     """
     photo = scipy.misc.imread(imgpath, flatten=True).astype(float) #Read as grayscale
     m, n = photo.shape
-    p = round(delta * (m * n - 1)) #Number of blanks
+    p = round(delta * (m * n - 1)) #Number of non-blanks
     A = np.zeros((m * n, 1), dtype=bool)
     A[0:p] = True
     ind = np.random.permutation(range(m*n))
@@ -19,7 +19,7 @@ def blur(imgpath, delta = 0.65):
 
 if __name__ == '__main__':
     
-    # Read image and randomly erase 65% of the pixels
+    # Read image and randomly erase 35% of the pixels
     photo, A = blur('./columbia_1.png')
     M = np.copy(photo)
     M[~A] = np.nan

### 2015-01-06 document/article.tex
update document

@@
 
-	\item Step Size: The algorithms performance is highly dependent on the step size (alpha) parameter of the update equations.
+	\item Step Size: The algorithms performance is highly dependent on the step size ($\alpha$) parameter of the update equations.
 Too small of a alpha will result in convergence taking a very long time and lower the chance of finding the global objective.
 Too large of a alpha will result in convergence not happening as the the algorithm continually 'overshoots' the optimal value.
+Thus an essential part of using this algorithm will be experimentation with the step size.
+Currently, the step size (alpha parameter) is fully adjustable, with the default being a .01 decreasing 30\% each pass over the data until it reaches 1e-6 whereby it will remain constant until the stopping criterion of the algorithm are reached.
 
-	\item Shuffle: When dealing with large data we need to be clever with how we shuffle our data.
-Shuffling is no trivial task out of memory.
-We explored several out of memory methods but ultimately decided with using a pseudo-shuffle, where we read in the file in large chunks that fit in memory, shuffle the lines using built in python functions, write the records back to a text file, and continue until the all of the file has been pseudo-shuffled.
-Hopefully by choosing a large enough batch size for reading in, our file is shuffle well enough.
-We also explored with a more robust method of shuffling the shuffled batches, but the added time complexity of appending large files together did not prove worthwhile given almost negligible returns in convergence.
+	\item Shuffle: When dealing with large data files we needed to be clever with how we shuffle our data.
+Stochastic gradient descent performs much better with shuffling of the data after each pass over the file thus shuffling is essential; yet shuffling is no trivial task for files too big to fit entirely into the computers random access memory.
+We explored various shuffling methods for both in-memory and out-of-memory but ultimately decided with using a batch pseudo-shuffle.
+Our batch shuffle method shuffles the file in chunks.
+So we read in some large portion (the batch size) of the file that will fit in memory, we shuffle the ratings of that batch using built in python functions, and lastly we write the shuffled ratings back to a text file, and continue sequentially until the all of the file has been shuffled.
+Following this method we say 'pseudo-shuffled' since the file is not truly shuffled.
+It is shuffled within the batch/portion of the file it belongs to but still not truly shuffled within the entire file.
+The larger the batches, the better the shuffled ratings, until of course we read the entire file into memory and shuffle it, the ideal situation.
+Hopefully by choosing a large enough batch size for reading in, our file will be shuffled well enough.
+We also explored a more robust method of shuffling the data by additionally shuffling the shuffled batches whereas before we wrote the shuffled ratings back to a file sequentially.
+However ultimately the noticeably bigger time complexity of moving and appending large files together via Linux did not prove worthwhile given almost negligible returns in convergence
 

### 2015-01-06 document/article.tex
yet more changes to the document

@@
 
-\title{\texttt{CompleteThat}: A python package for  low-rank matrix completion\\EEOR E4650 Course Project
+\title{\texttt{CompleteThat}\\ A python package for  low-rank matrix completion\\EEOR E4650 Course Project
 }
@@
 Given a low rank matrix with partial entries we implemented different methods to solve the matrix completion problem:  First, for ``small'' problems we implemented two memory-based algorithms  following the work by Tanner and Wei~\cite{Tanner:2014} to find, given a number $r$, the matrix with rank $r$ that best fits the data using the Frobenius norm.
-Second, when the problem does not fit into memory, we implemented a memory-fitting algorithm (stochastic gradient descent) following the approach in Zhang~\cite{zhang:2004} and Bottou~\cite{bottou:2012}.
+Second, when the problem does not fit into memory, we implemented an out-of-core algorithm (stochastic gradient descent) following the approach in Zhang~\cite{zhang:2004} and Bottou~\cite{bottou:2012}.
 The package is available at \url{https://pypi.python.org/pypi/completethat/0.1dev}.
@@
 \subsubsection*{Low rank matrix completion by stochastic gradient descent}
-The formal problem as defined in the model section is stated in the language of linear algebra and the typical optimizations using gradient methods require access to all of the data at once for updating the parameters, since its dependent on calculating inverse of matrices or solving systems of linear equations.
-However, given the scale of data that is common in today's digital world, the need for more efficient and memory friendly algorithms arises.
+The formal problem as defined in the model section is stated in the standard language of linear algebra and the typical optimizations using gradient methods require the calculation of the gradient based on the sum of all the error terms, meaning that when updating our latent feature vectors, we calculate the gradient based on all the data in our ratings file.
+Given the scale of data that is common in today's digital world, the need for more efficient and memory friendly algorithms arises as calculating the gradient on a large enough file will be computationally expensive and time consuming.
 
@@
 
-With stochastic gradient descent, a variation of the standard batch gradient optimization, instead of calculating the gradient for the user feature vector using all the item feature vectors that the user has rated as would be done  usually, we approximate the user's gradient using the current item vector for the current rating and vice versa for updates of the item vectors.
-Thus we are essentially exchanging the luxury of using less memory and computational resources for more time and iterations spent with stochastic gradient descent.\footnote{We note that our formulation is not convex but, due to shuffling of the data and the way the gradients are estimated, the algorithm will converge to at least a local minimum.}
+With Stochastic gradient descent, a variation of the standard batch gradient optimization, instead of calculating the gradient using all of the ratings data, we approximate the gradient of our objective function using just the current record and update just the latent feature vectors of the user and item pertaining to the current rating.
+Thus one of the highlights of our package is that we simply read the file in line by line, running the algorithm in place.
+After each update, we simply read in the next line of the file and repeat.\footnote{We note that our formulation is not convex but, due to shuffling of the data and the way the gradients are estimated, the algorithm will converge to at least a local minimum.}
+
+Our RAM requirements for the algorithm decrease significantly with this sort of out-of-memory implementation.
+For example, with the Yahoo Music Data we only needed to store the parameters for each user and item, rather than the entire 115 million records of our file plus the parameters.
+This results in storing ~2 million parameters in a python dictionary (a sort of built in hash function and table) rather than 115 million records in memory.
+Thus we are essentially exchanging the luxury of using less memory and computational resources for more time and iterations spent with stochastic gradient descent.
+
+After the algorithm converges or the maximum number of iterations are me have a set of user and item latent vectors which can be used to compute the estimated rating $r_{ij}$ by simply taking the dot product of the two vectors.
+Similarly the entire ratings matrix could easily be built with the vectors although we have neglected to do so for practicality reasons.
 
@@
 \subsubsection*{MatrixCompletionBD}
-Given a csv file with the input data, the matrix completion problem can be solved as:
+Given a csv file with the input data with three fields (user id, item id , rating) the matrix completion problem can be solved as:
 
@@
 The Yahoo music and movie data sets are publicly available datasets published by Yahoo for use to the public for recreational and research purposes.
-The music training data set is a small 4 mb pipe-delimited file consisting of roughly 210,000 user-movie ratings.
+The music training data set is a small 4 mb tab-delimited file consisting of roughly 210,000 user-movie ratings.
 The test set has roughly 10,000 records.
-There are two ratings schemas available, one on a 5-point scale and the other on a 13-point scale and we arbitrarily chose to use the 5-point scale.
+There are actually two ratings schemas available, one on a 5-point scale and the other on a 13-point scale and we arbitrarily chose to use the 5-point scale.
 The Yahoo music data is a larger 2.2 gb file of 115 million user-song ratings.
@@
 
-We look to evaluate the performance of our SGD algorithm on these two datasets comparing it versus a naive benchmark estimate, where we used the mean of the training set as our estimator for all records in the test set.
-Then we compared how well the algorithms performed versus the benchmark.
+We look to evaluate the performance of our algorithms on these two datasets, comparing the various algorithms implemented versus a naive benchmark estimate where we used the mean of the training set as our estimator for all records in the test set.
+Then we compared how well the algorithms performed versus the benchmark as well as Mahout.
+Apache Mahout is a mainstream scalable and fast machine learning library implemented in java which has a recommendation engine component similar to MatrixCompletionBD but uses alternating least squares for the optimization.
+Thus given the origins and popularity of Mahout, we are also interested in comparing it to our python package?s performance.
+The naïve estimator for the yahoo music data is 4.088.
+This is the mean value of the ratings in the training set so we will compute the MSE on the test set using the benchmark, which yields an RMSE of 1.10 (Note we use RMSE here only because Mahout uses RMSE).
+A quick run of Mahout yields 1.17 and CompleteThat's SGD model yielded a 1.19 (we used rank 7 but larger ranks will outperform the naïve baseline as we show in our table).
+So our model yielded a pretty similar RMSE as the MAHOUT package.
+Note that for both of these instances we used a factorization of rank equal to seven and the CompleteThat stochastic gradient descent algorithm ran slightly faster than Mahout's, with our SGD algorithm running on 220,000 ratings and converging in less than a minute (.537 minutes versus 1.4 minutes)!
 
-The benchmark for the yahoo music data is 4.088.
-This is the mean value of the ratings in the training set so we will compute the MSE on the test set using the benchmark which yields an RMSE of 1.10.
-A quick run of Mahout yields 1.17 and CompleteThat's sgd model yielded a 1.19.
-So our model yielded a very similar RMSE as the MAHOUT package.
-Note that for both of these instances we used a factorization of rank equal to seven and the Complete that sgd algorithm ran slightly faster than Mahout's (.537 minutes vs 1.4 minutes).
+The Yahoo Music data is a lot larger and provided a much more interesting scenario but unfortunately due to the time constraints and how long it took to converge on the Yahoo Music data, we didn?t get as much testing done here.
+It does take a very long time for the stochastic gradient descent algorithm to converge, but does run.
+In experimentation with features of rank six, we saw it converging in approximately 9 hours, with results roughly on par relative to the naïve baseline using the training set.
+Unfortunately, we weren?t able to get Apache Mahout working on our local machines with that scale of data, thus have no real comparison to be made in terms of how well our package does scale, but if nothing else we can say that our algorithm is fairly invariant to the size of the data.
 
+Table~\ref{tab:yahoo_data} shows the performance of the SGD on the Yahoo music with varying the rank of the user and item latent features.
+We see that the algorithm performs best on the test set using a rank around 15, and it is evident that too large a rank is detrimental to performance.
 \begin{table}[h]
@@
 \caption{RMSE vs Step Size vs Rank}
+\label{tab:yahoo_data}
 \end{table}
@@
 Overall, we would like to test the software and integrate automated testing cases into the software developing cycle.
-Also, documentation for the package and its functions, including examples, is high in our priority list.
+Also, documentation for the package and its functions, including examples, is high on our priority list.
 In addition, we hope to incorporate more rigorous error checking and make   available more customization to the user.

### 2015-01-06 document/article.tex
yet more changes to the document

@@
 
-where $\sigma_i(X)$ denotes the $i$th largest singular value of $X$ and $\|X\|_*$ is the called the nuclear norm.
+where $\sigma_i(X)$ denotes the $i$th largest singular value of $X$ and $\|X\|_*$ is called the nuclear norm.
 The main point of this relaxation is that the nuclear norm is a convex function and thus can be optimized efficiently via semidefinite programming or by iterative soft thresholding algorithms~\cite{cai:2010}~\cite{goldfarb:2011}.
 
-Alternative to nuclear norm minimization there have been many algorithms which are designed to target the following optimization problem
+As an alternative to the nuclear norm minimization, there exists several algorithms designed to solve the following optimization problem
 \begin{equation}
@@
 where $X=YZ$ is the completed matrix and $Y\in \reals^{n_1\times r}, Z\in \reals^{r\times n_2}$ and $r$ represents the (hopefully small) rank of the matrix $X$.
-The main point of this relaxation is the use of an alternating minimization scheme or Gauss-Seidel or 2-block coordinate descent method, where first $Y$ is fixed and then the problem is reduced to a convex, standard least squares problem on $Z$.
-Later, $Z$ is fixed and then the problem is reduced to a convex, standard least squares problem on $Y$.
+The main point of this relaxation is the use of an alternating minimization scheme or Gauss-Seidel or 2-block coordinate descent method, where first $Y$ is fixed and then the problem is reduced to a convex, standard least squares problem, on $Z$.
+Later, $Z$ is fixed and then the problem is reduced to a convex, standard least squares problem, on $Y$.
 
@@
 We allow the user to choose between several optimization algorithms including two in memory based algorithms, alternating steepest descent and scaled alternating steepest descent, and one out of memory fitting procedure using stochastic gradient descent.
-We use extensively the numerical libraries \texttt{spicy} and \texttt{numpy}, and the current implementation includes two classes, \texttt{MatrixCompletion} for in memory based algorithms and \texttt{MatrixCompletionBD} for the memory fitting procedure.
+We use extensively the numerical libraries \texttt{spicy} and \texttt{numpy}, and the current implementation includes two classes, \texttt{MatrixCompletion} for in memory based algorithms and \texttt{MatrixCompletionBD} for the out-of-core procedure.
 The package is available worldwide and can be
@@
 
-Our RAM requirements for the algorithm decrease significantly with this sort of out-of-memory implementation.
-For example, with the Yahoo Music Data we only needed to store the parameters for each user and item, rather than the entire 115 million records of our file plus the parameters.
-This results in storing ~2 million parameters in a python dictionary (a sort of built in hash function and table) rather than 115 million records in memory.
-Thus we are essentially exchanging the luxury of using less memory and computational resources for more time and iterations spent with stochastic gradient descent.
-
-After the algorithm converges or the maximum number of iterations are me have a set of user and item latent vectors which can be used to compute the estimated rating $r_{ij}$ by simply taking the dot product of the two vectors.
-Similarly the entire ratings matrix could easily be built with the vectors although we have neglected to do so for practicality reasons.
+After the algorithm converges or the maximum number of iterations are met, the algorithm has a set of user and item latent vectors which can be used to compute the estimated rating $r_{ij}$ by simply taking the dot product of the two vectors.
+Similarly, the entire ratings matrix could easily be built with the vectors although we have neglected to do so for practicality reasons.
 
@@
 Then we compared how well the algorithms performed versus the benchmark as well as Mahout.
+
 Apache Mahout is a mainstream scalable and fast machine learning library implemented in java which has a recommendation engine component similar to MatrixCompletionBD but uses alternating least squares for the optimization.
-Thus given the origins and popularity of Mahout, we are also interested in comparing it to our python package?s performance.
-The naïve estimator for the yahoo music data is 4.088.
+Thus given the origins and popularity of Mahout, we are also interested in comparing it to our python packages performance.
+
+The naive estimator for the yahoo music data is 4.088.
 This is the mean value of the ratings in the training set so we will compute the MSE on the test set using the benchmark, which yields an RMSE of 1.10 (Note we use RMSE here only because Mahout uses RMSE).
-A quick run of Mahout yields 1.17 and CompleteThat's SGD model yielded a 1.19 (we used rank 7 but larger ranks will outperform the naïve baseline as we show in our table).
+A quick run of Mahout yields 1.17 and CompleteThat's SGD model yielded a 1.19 (we used rank 7 but larger ranks will outperform the naive baseline as we show in our table).
 So our model yielded a pretty similar RMSE as the MAHOUT package.
@@
 
-The Yahoo Music data is a lot larger and provided a much more interesting scenario but unfortunately due to the time constraints and how long it took to converge on the Yahoo Music data, we didn?t get as much testing done here.
+The Yahoo Music data is a lot larger and provided a much more interesting scenario but unfortunately due to the time constraints and how long it took to converge on the Yahoo Music data, we did not did as much testing.
 It does take a very long time for the stochastic gradient descent algorithm to converge, but does run.
 In experimentation with features of rank six, we saw it converging in approximately 9 hours, with results roughly on par relative to the naïve baseline using the training set.
-Unfortunately, we weren?t able to get Apache Mahout working on our local machines with that scale of data, thus have no real comparison to be made in terms of how well our package does scale, but if nothing else we can say that our algorithm is fairly invariant to the size of the data.
+Unfortunately, we were not able to get Apache Mahout working on our local machines with that scale of data, thus have no real comparison to be made in terms of how well our package does scale, but if nothing else we can say that our algorithm is fairly invariant to the size of the data.
+
+Our RAM requirements for the algorithm decrease significantly with this sort of out-of-memory implementation.
+For example, with the Yahoo Music Data we only needed to store the parameters for each user and item, rather than the entire 115 million records of our file plus the parameters.
+This results in storing ~2 million parameters in a python dictionary (a built-in hash table) rather than 115 million records in memory.
+Thus we are essentially exchanging the luxury of using less memory and computational resources for more time and iterations spent with stochastic gradient descent.
 

### 2015-01-06 document/article.tex
Fixed Grammatical Errors

@@
 For our convex optimization project, we decided to take some of the more interesting and applicable topics and algorithms of our class and develop a python package to implement them.
-We briefly discuss the theory behind matrix factorizing, the models and approaches we choose to use and the algorithms for the optimization.
+We briefly discuss the theory behind matrix factorization, the models and approaches we chose to use, and the algorithms for the optimization.
 Later, we walk through two case studies illustrating the usefulness and applicability in the context of image processing and a recommendation system for Yahoo music and movie data.
@@
 \section{Matrix Completion Problem}
-The matrix completion problem can be formally stated as follows.
-We are interested in recovering a matrix $M \in \reals^{n_1 \times n_2}$ but only get to observe a number $m \ll n_1n_2$ of its entries.
-Thus, we want to find a solution to the following optimization problem
+The matrix completion problem can be formally described as follows: we are interested in recovering a matrix $M \in \reals^{n_1 \times n_2}$ but only get to observe a number $m \ll n_1n_2$ of its entries.
+Thus, we want to find a solution to the following optimization problem:
 
@@
 where $X\in \reals^{n_1 \times n_2}$ is the decision variable and $\rank(X)$ is equal to the rank of the matrix $X$.
-The problem (\ref{eq:original_problem1}) seeks the simplest matrix fitting the observed data.
+The defined problem (\ref{eq:original_problem1}) seeks the simplest matrix fitting the observed data.
 Of course, if there were only one low-rank object fitting the data, this would recover $M$.
@@
 
-Unfortunately, this optimization problem has been shown to be NP-hard and all know algorithms which provide exact solutions require time doubly exponential in the dimension of the matrix both in theory and in practice~\cite{Candes:2009}.
+Unfortunately, this optimization problem has been shown to be NP-hard and all known algorithms which provide exact solutions require exponential time complexity in the dimension of the matrix both in theory and in practice~\cite{Candes:2009}.
 Also, the $\rank(X)$ makes the problem non-convex.
@@
 Given a low rank matrix with partial entries the package solves an optimization problem to estimate the missing entries.
-We allow the user to choose between several optimization algorithms including two in memory based algorithms, alternating steepest descent and scaled alternating steepest descent, and one out of memory fitting procedure using stochastic gradient descent.
-We use extensively the numerical libraries \texttt{spicy} and \texttt{numpy}, and the current implementation includes two classes, \texttt{MatrixCompletion} for in memory based algorithms and \texttt{MatrixCompletionBD} for the out-of-core procedure.
+We allow the user to choose between several optimization algorithms including two in-memory based algorithms, alternating steepest descent and scaled alternating steepest descent, and one out-of-core fitting procedure using stochastic gradient descent.
+We utilize extensively the python numerical libraries \texttt{scipy} and \texttt{numpy}, and the current implementation includes two classes, \texttt{MatrixCompletion} for in memory based algorithms and \texttt{MatrixCompletionBD} for the out-of-core procedure.
 The package is available worldwide and can be
@@
 
-Where M is a MATLAB matrix with nan on the missing entries and \texttt{X(index) == M(index)} corresponds to the $P_{\Omega}(X) = P_{\Omega}(M)$ restriction.
+Where M is a MATLAB matrix with 'nan' on the missing entries and \texttt{X(index) == M(index)} corresponds to the $P_{\Omega}(X) = P_{\Omega}(M)$ restriction.
 However, the computation is very slow and for any matrix greater than 100x100 the computational time is too high.
-Since this is clearly unsatisfactory for practical purposes, specific algorithms were implemented on the package.
+Since this is clearly unsatisfactory for practical purposes, specific algorithms were implemented in this package to handle matrices on the scale more commonly seen in industry.
 
@@
 \subsubsection*{Low rank matrix completion by alternating steepest descent methods}
-The  alternating steepest descent methods implemented on \texttt{CompleteThat} solve the Problem (\ref{eq:frob_problem}) above.
+The  alternating steepest descent methods implemented in \texttt{CompleteThat} solves the Problem (\ref{eq:frob_problem}) above.
 Let the objective function of the optimization problem $f: \reals^{n_1\times r}\times \reals^{r\times n_2} \rightarrow \reals$ be defined as
@@
 
-To solve efficiently the above problem, Tanner and Wei~\cite{Tanner:2014} use a single step of simple line-search along the gradient descent directions.
+To efficiently solve the above problem, Tanner and Wei~\cite{Tanner:2014} use a single-step, simple line-search along the gradient descent directions.
 Let us write $f(Y, Z)$ as $f_Z(Y)$ when $Z$ is held constant and $f_Y(Z)$ when $Y$ is held constant and let $t_y, t_z$ be the steepest descent step sizes for descent directions.
@@
 The Scaled Alternating Steepest Descent Method (sASD) is an accelerated version of ASD and uses a scaled gradient descent direction with exact line-search.
-Its main motivation for the scaled factors is the explicit expression for the Newton directions for the problem (\ref{eq:ASD_prob}) if all the entries in the matrix $M$ were know:
+Its main motivation for the scaled factors is the explicit expression for the Newton directions for the problem (\ref{eq:ASD_prob}) if all the entries in the matrix $M$ were known:
 \[
@@
     \mbox{minimize}_{r_{ij}\in K}   & f:= (r_{ij} - \hat r_{ij})^2 = e_{ij}^2\\
-    \mbox{subject to} & r_{ij} = p_i^T q_j
+    \mbox{subject to} & \hat r_{ij} = p_i^T q_j
     \end{array}
@@
 After the algorithm converges or the maximum number of iterations are met, the algorithm has a set of user and item latent vectors which can be used to compute the estimated rating $r_{ij}$ by simply taking the dot product of the two vectors.
-Similarly, the entire ratings matrix could easily be built with the vectors although we have neglected to do so for practicality reasons.
+Similarly, the entire ratings matrix could easily be built with the vectors although we have neglected to do so in our package for practicality reasons.
 
@@
 Thus an essential part of using this algorithm will be experimentation with the step size.
-Currently, the step size (alpha parameter) is fully adjustable, with the default being a .01 decreasing 30\% each pass over the data until it reaches 1e-6 whereby it will remain constant until the stopping criterion of the algorithm are reached.
+Currently, the step size (alpha parameter) is fully adjustable, with the default being a .01 decreasing 30\% each pass over the data until it reaches 1e-6 whereby it will remain constant until the stopping criterion of the algorithm is reached.
 
@@
 Our batch shuffle method shuffles the file in chunks.
-So we read in some large portion (the batch size) of the file that will fit in memory, we shuffle the ratings of that batch using built in python functions, and lastly we write the shuffled ratings back to a text file, and continue sequentially until the all of the file has been shuffled.
+So we read in some large portion (the batch size) of the file that will fit in memory, we shuffle the ratings of that batch using built in python functions, we write the shuffled ratings back to a text file, and continue sequentially until the all of the file has been shuffled.
 Following this method we say 'pseudo-shuffled' since the file is not truly shuffled.
@@
 The larger the batches, the better the shuffled ratings, until of course we read the entire file into memory and shuffle it, the ideal situation.
-Hopefully by choosing a large enough batch size for reading in, our file will be shuffled well enough.
+Hopefully by choosing a large enough batch size for reading into the memory, our file will be shuffled well enough.
 We also explored a more robust method of shuffling the data by additionally shuffling the shuffled batches whereas before we wrote the shuffled ratings back to a file sequentially.
-However ultimately the noticeably bigger time complexity of moving and appending large files together via Linux did not prove worthwhile given almost negligible returns in convergence
+However ultimately the noticeably bigger time complexity of moving and appending large files together via Linux did not prove worthwhile given almost negligible returns in convergence.
 
@@
 Apache Mahout is a mainstream scalable and fast machine learning library implemented in java which has a recommendation engine component similar to MatrixCompletionBD but uses alternating least squares for the optimization.
-Thus given the origins and popularity of Mahout, we are also interested in comparing it to our python packages performance.
+Thus given the origins and popularity of Mahout, we are also interested in comparing it to our python package's performance.
 
-The naive estimator for the yahoo music data is 4.088.
-This is the mean value of the ratings in the training set so we will compute the MSE on the test set using the benchmark, which yields an RMSE of 1.10 (Note we use RMSE here only because Mahout uses RMSE).
+The naive estimator for the Yahoo music data is 4.088.
+This is the mean value of the ratings in the training set so we will compute the MSE on the test set using the benchmark, which yields an RMSE of 1.10 (Note we switch to RMSE here only because Mahout uses RMSE).
 A quick run of Mahout yields 1.17 and CompleteThat's SGD model yielded a 1.19 (we used rank 7 but larger ranks will outperform the naive baseline as we show in our table).
-So our model yielded a pretty similar RMSE as the MAHOUT package.
-Note that for both of these instances we used a factorization of rank equal to seven and the CompleteThat stochastic gradient descent algorithm ran slightly faster than Mahout's, with our SGD algorithm running on 220,000 ratings and converging in less than a minute (.537 minutes versus 1.4 minutes)!
+So our model yielded a pretty similar RMSE as the MAHOUT package although both were worse than the benchmark.
+Later with some fine-tuning we beat the benchmark.
+*Note that for both of these instances we used a factorization of rank equal to seven and the CompleteThat stochastic gradient descent algorithm ran slightly faster than Mahout's, with our SGD algorithm running on 220,000 ratings and converging in less than a minute (.537 minutes versus 1.4 minutes)!
 
-The Yahoo Music data is a lot larger and provided a much more interesting scenario but unfortunately due to the time constraints and how long it took to converge on the Yahoo Music data, we did not did as much testing.
+The Yahoo Music data is a lot larger and provided a much more interesting scenario but unfortunately due to the time constraints and how long it took to converge on the Yahoo Music data, we did not do as much testing.
 It does take a very long time for the stochastic gradient descent algorithm to converge, but does run.
-In experimentation with features of rank six, we saw it converging in approximately 9 hours, with results roughly on par relative to the naïve baseline using the training set.
+In experimentation with latent feature vectors of length six, we saw it converging in approximately 9 hours, with results roughly on par relative to the naïve baseline using the training set.
 Unfortunately, we were not able to get Apache Mahout working on our local machines with that scale of data, thus have no real comparison to be made in terms of how well our package does scale, but if nothing else we can say that our algorithm is fairly invariant to the size of the data.
@@
 In this section, we demonstrate the applicability of the package to image processing problems.
-Grayscale pictures can be transform into a rectangular matrix where each element of the matrix determines the intensity of the corresponding pixel.
+Grayscale pictures can be transformed into a rectangular matrix where each element of the matrix determines the intensity of the corresponding pixel.
 For convenience, most of the current digital files use integer numbers between 0 and 255.
@@
 For illustration we used three  512 x 512 grayscale photographs taken by one of the authors of the Columbia University campus on December, 2014.
-The original, erased and reconstructed images are shown on Figure~\ref{fig:columbia_pics}, where we erased 35\% of the pixels of each image.
+The original, erased, and reconstructed images are shown on Figure~\ref{fig:columbia_pics}, where we erased 35\% of the pixels of each image.
 In addition, to illustrate the ease of use of the package, we also present the (very short) script used to recover the photographs:
@@
 We have plans for various improvements as we continue working with the \texttt{CompleteThat} python package.
-Overall, we would like to test the software and integrate automated testing cases into the software developing cycle.
+Overall, we would like to test the software and integrate automated testing cases into the software development cycle.
 Also, documentation for the package and its functions, including examples, is high on our priority list.
-In addition, we hope to incorporate more rigorous error checking and make   available more customization to the user.
+In addition, we hope to incorporate more rigorous error checking and make more customization available to the user.
 
-For the memory-based algorithms it is well know that noise in the data introduces overfitting on the estimated matrix.
-Following the approach taken by Mazumder et al.~\cite{mazumder:2010}  where they solve the same objetive function as the problems above (using the Frobenious norm) but introducing the nuclear norm as regularizer to account for overfitting, we would like to introduce a regularization option into the matrix completion procedure.
+For the memory-based algorithms it is well known that noise in the data introduces overfitting on the estimated matrix.
+Following the approach taken by Mazumder et al.~\cite{mazumder:2010}  where they solve the same objective function as the problems above (using the Frobenious norm) but introducing the nuclear norm as regularizer to account for overfitting, we would like to introduce a regularization option into the matrix completion procedure.
 
